---
title: "watsi's patient data exploration"
output: html_document
---

This is an analysis of patient data from watsi.org. The link to data-set is : [Watsi data](https://docs.google.com/spreadsheets/d/1tZq47h6jg7NX4ddhTS_H8JFVfLZiDbxwwdQD47_ow64/edit "Watsi dataset")


watsi.org brings kickstarter model to health needs of individuals from developing countries. Users can fund a patient, once a target for the funds needed to be raised is reached, the funds gets transferred for treating patients. 100% of proceeds from users go directly to the patient's fund. 

We are going to look at the patient data and explore various parameters in this data-set. 

```{r, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)

patientDF <- read.csv("Patients.csv", sep = ',')

# plot text adjustments
eleText <- element_text(size=10)
```


Load the data set. The following are the columns in the data set:
```{r, echo=FALSE}
names(patientDF)
```

To Analyze the data, we will add two more columns to the data set - "yearFunded" & "monthFunded".
```{r, echo=FALSE}
patientDF$yearFunded <- as.numeric(strftime(as.Date(patientDF$Date.Funded), format = "%Y"))
patientDF$monthFunded <- as.numeric(strftime(as.Date(patientDF$Date.Funded), format = "%m"))
patientDF$monthFundedWord <- strftime(as.Date(patientDF$Date.Funded), format = "%b")
```

Let us take a look at the cost distribution. Following are two histograms of cost with bin-width 500, 50 respectively. As we see majority of the "Cost" values are in range 0-500. i.e., most patient funding needed is less than $500.
```{r, echo=FALSE}
patientDF$Cost <- as.numeric(gsub("[$,]","", patientDF$Cost))
ggplot(data = patientDF, aes(x=Cost)) +
  geom_histogram(binwidth = 500) +
  scale_x_continuous(breaks = seq(0, max(patientDF$Cost), by=500)) +
  labs(title = "Histogram of Cost. binwidth = 500") +
  ylab("Patient count") +
  xlab("Cost in $") + 
  theme(plot.title = eleText, axis.title = eleText)

ggplot(data = patientDF, aes(x=Cost)) +
  geom_histogram(binwidth = 50) +
  scale_x_continuous(breaks = seq(0, max(patientDF$Cost), by=250)) +
  labs(title = "Histogram of Cost. binwidth = 50") +
  ylab("Patient count") +
  xlab("Cost in $") +
  theme(plot.title = eleText, axis.title = eleText)
```

In fact, as we see below the median is $512 and the mean is $604.5
```{r, echo=FALSE}
summary(patientDF$Cost)
```

Now, let us take a look at the countries which patients belong to. As we see below, Cambodia has the highest number of patient, followed by Kenya and Tanzania. 

```{r, echo=FALSE}
pByCountry <- group_by(patientDF, Country)
pCount_byCountry <- summarise(pByCountry, count = n())

pCount_byCountry$Country <- factor(pCount_byCountry$Country, levels = pCount_byCountry$Country[order(pCount_byCountry$count)])

ggplot(data = pCount_byCountry, aes(x = Country, y = count)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Patient Count in Countries") +
  ylab("Patient count") +
  theme(plot.title = eleText, axis.title = eleText)
```

Now let's take a look at the costs per country. The total costs funded per country closely resembles the previous graph, but Cambodia no longer has the highest total costs. This perhaps might be related to lower medical cost in Cambodia?

```{r, echo=FALSE}
cost_byCountry <- summarise(pByCountry, costCount = sum(Cost))

cost_byCountry$Country <- factor(cost_byCountry$Country, levels = cost_byCountry$Country[order(cost_byCountry$costCount)])

ggplot(data = cost_byCountry, aes(x = Country, y = costCount)) +
  geom_bar(stat= "identity") +
  coord_flip() +
  scale_y_continuous(ylab("Total Cost in $1000"), breaks = seq(0, 800000, by=100000), labels = seq(0, 800, by = 100)) +
  labs(title = "Total cost per country") +
  theme(plot.title = eleText, axis.title = eleText)
```

A look at mean costs by country, it seems to be the case that mean medical cost needed are indeed lowest in Cambodia. However, we should be cautious to claim that outright, since there could be other factors affecting "cost" in Cambodia, but there seems to be some indication towards that. 

```{r, echo=FALSE}
meancost_byCountry <- summarise(pByCountry, costCount = mean(Cost))

meancost_byCountry$Country <- factor(meancost_byCountry$Country, levels = meancost_byCountry$Country[order(meancost_byCountry$costCount)])

ggplot(data = meancost_byCountry, aes(x = Country, y = costCount)) +
  geom_bar(stat= "identity") +
  coord_flip() +
  scale_y_continuous(ylab("Mean Cost in $")) +
  labs(title = "Mean cost per country") +
  theme(plot.title = eleText, axis.title = eleText)
```

Number of days it took to fund patients. Using the "Date.Posted" and "Date.Funded" we can calculate the number of days it took to fund each patient, we call this column "numDaysFunded". Let us take a look at how "numDaysFunded" changes for various factors. 

Following is a summary of number of days it took to fund patients. It is a bit surprising and perhaps impressive that on average it took just 3.271 days to fund patients.
```{r, echo=FALSE}
patientDF$numDaysFunded <- as.numeric(as.Date(patientDF$Date.Funded) - as.Date(patientDF$Date.Posted))
#nonNADaysFundedData = patientDF[!is.na(patientDF$numDaysFunded),]
nonNADaysFundedData = subset(patientDF, !is.na(patientDF$numDaysFunded))
summary(nonNADaysFundedData$numDaysFunded)

```


```{r, echo=FALSE, warning=FALSE, error=FALSE}
meanNumDays = mean(nonNADaysFundedData$numDaysFunded);
ggplot(data = nonNADaysFundedData, aes(x=numDaysFunded)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 30), xlab("Number of days for funding"), breaks = seq(0, 30, by=5)) +
  geom_vline(xintercept = meanNumDays, color = "blue") +
  geom_text(aes(x=meanNumDays, label="mean = 3.271", y= 1000), colour="blue", angle=90, vjust = -0.4, fontface = 'plain') +
  labs(title = "Histogram of Number of Days for successfully funding a patient. Binwidth = 1 day") + 
  theme(plot.title = eleText, axis.title = eleText)
```


```{r, echo=FALSE, eval= FALSE}
ggplot(data = patientDF, aes(x=monthFunded)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(labels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"), breaks = seq(1.5, 12.5, by = 1), limits = c(1, 13) ) +
```

```{r, echo=FALSE, warning=FALSE}
daysFunded_byMonth <- group_by(nonNADaysFundedData, monthFunded, monthFundedWord)
meanNumDaysFunded_byMonth <- summarise(arrange(daysFunded_byMonth, monthFunded), meanDaysFunded = mean(numDaysFunded))

# fix levels from Jan - Dec (instead of alphabetical order)
nonNADaysFundedData$monthFundedWord <- factor(nonNADaysFundedData$monthFundedWord, levels = meanNumDaysFunded_byMonth$monthFundedWord)

ggplot(data = nonNADaysFundedData, aes(x=numDaysFunded)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(limits = c(0, 30), xlab("Number of days for funding"), breaks = seq(0, 30, by=5)) +
  facet_wrap(~monthFundedWord) +
  labs(title = "Histogram of Number of Days for successfully funding a patient grouped by Month. Binwidth = 1 day.") +
  theme(plot.title = eleText, axis.title = eleText)

#geom_vline(data = meanNumDaysFunded_byMonth, aes(xintercept=meanDaysFunded), color = "blue") +
#  geom_text(data = meanNumDaysFunded_byMonth, aes(x=meanDaysFunded, label=round(meanDaysFunded, digits = 1), y= 275), colour="blue", angle=90, vjust = 1.1, fontface = 'plain') +
```

Following is a consolidated view of all patients funded during Watsi's existent, grouped by month. December seems to be most philanthropic month, which is not too surprising considering December is the biggest month for charities and other non-profit organizations when it comes to fund raising. 

```{r, echo=FALSE}
patientDF$monthFunded <- as.numeric(strftime(as.Date(patientDF$Date.Funded), format = "%m"))

pbyMonthFunded <- group_by(patientDF, monthFunded)
pcountByMonthFunded <- summarise(pbyMonthFunded, count = n())
# drop na month. Since the count is very small for na(12). safe to do 
pcountByMonthFunded <- pcountByMonthFunded[!is.na(pcountByMonthFunded$monthFunded),]
ggplot(data = pcountByMonthFunded, aes(x=monthFunded, y = count)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(labels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"), breaks = seq(1, 12, by = 1), limits = c(0, 13) ) +
  labs(title = "Number of patients funded per month. Data from 2012-2016.") +
  xlab("Month") +
  ylab("Count") +
  theme(plot.title = eleText, axis.title = eleText)
```

View of the same chart as above but grouped by year. 

```{r, eval = FALSE, echo=FALSE}
ggplot(data = patientDF, aes(x=monthFunded)) +
  geom_histogram(binwidth = 1) +
  scale_x_continuous(labels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"), breaks = seq(1.5, 12.5, by = 1), limits = c(1, 13) ) +
  facet_wrap(~yearFunded, ncol = 1) +
  ylab("Patients Funded") +
  xlab("Month") +
  labs(title = "Patients funded per month from 2012-2016") +
  theme(plot.title = eleText, axis.title = eleText)
```

Considering that the year just started, looks like Watsi has got to a good start, especially considering the uptick in number of patients funded in February compared to previous year.  
```{r, echo=FALSE}
nonNAMonthsData <- patientDF[!is.na(patientDF$monthFunded), ]
ggplot(data = nonNAMonthsData, aes(x=monthFunded)) +
  geom_bar() +
  scale_x_continuous(labels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec"), breaks = seq(1, 12, by = 1), limits = c(0, 13) ) +
  facet_wrap(~yearFunded, ncol = 1) +
  ylab("Patients Funded") +
  xlab("Month") +
  labs(title = "Patients funded per month from 2012-2016") +
  theme(plot.title = eleText, axis.title = eleText)
```

The upward trend of number of patients funded year-over-year is clear in the following chart.

```{r, echo= FALSE, warning=FALSE}
pdf_byYearFunded <- group_by(patientDF, yearFunded)
pbyYearFunded <- summarise(pdf_byYearFunded, count = n())
# There are some NA values here (12), which are excluded
ggplot(data = pbyYearFunded, aes(x = yearFunded, y = count)) +
  geom_bar(stat = "identity") +
  xlab("Year") +
  ylab("Patient Count") +
  labs(title = "Patient successfully funded per year") +
  theme(plot.title = eleText, axis.title = eleText)
```
